Neural Connectome of the Ctenophore Statocyst

### 2025-05-28 manuscript/Manuscript.tex

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\title{Neural Connectome of the Ctenophore Statocyst}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\textbf{Kei Jokura}\(^{1,2,3,4,5*}\), \textbf{Sanja
Jasek}\(^{1,2,6}\), \textbf{Lara Kewalow}\(^{6}\), \textbf{Pawel
Burkhardt}\(^{7}\), \textbf{Gáspár Jékely}\(^{1,2,6,*}\) \(^{1}\) Living
Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom
\(^{2}\) Biosciences, Faculty of Health and Life Sciences, University of
Exeter, Exeter EX4 4QD, UK \(^{3}\) Grass Laboratory, Marine Biological
Laboratory, Woods Hole, MA 02543, USA \(^{4}\) Exploratory Research
Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan
\(^{5}\) National Institute for Basic Biology (NIBB), Okazaki, 444-8585,
Japan \(^{6}\) Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany \(^{7}\) Michael Sars Centre,
University of Bergen, Norway \textbf{\(^{*}\) Correspondence:}
\href{mailto:jokura@nibb.ac.jp}{\nolinkurl{jokura@nibb.ac.jp}},
\href{mailto:gaspar.jekely@cos.uni-heidelberg.de}{\nolinkurl{gaspar.jekely@cos.uni-heidelberg.de}}}
\author{}
\date{}

@@

Ctenophores possess a unique gravity receptor (statocyst) in their
aboral organ formed by four clusters of ciliated balancer cells that
collectively support a statolyth. During reorientation, differential
load on the balancer cilia leads to altered beating of the ciliated comb
rows to elicit turns. To study the neural bases of gravity sensing, we
imaged by volume electron microscopy (vEM) the aboral organ of the
ctenophore \emph{Mnemiopsis leidyi}. We reconstructed 909 cells,
including syncytial neurons that form a nerve net. The syncytial neurons
synapse on the balancer cells and also form reciprocal connection with
the bridge cells that span the statocyst. High-speed imaging revealed
that balancer cilia beat and arrest in a coordinated manner but with
differences between the sagittal and tentacular planes of the animal,
reflecting nerve-net organisation. Our results suggest a coordinating
rather than effector function for the nerve net and inform our
understanding of the diversity of nervous-systems organisation across
animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that
actively swim by beating rows of fused cilia known as comb plates. These
coordinated ciliary movements not only generate propulsion but also
allow precise control of body orientation and directional changes within
the water column. Despite lacking a centralized nervous system,
ctenophores exhibit sophisticated and dynamic behavioral patterns and
respond to a broad range of external stimuli, suggesting the presence of
an evolutionarily unique and functionally integrated sensorimotor
system. Several components of the ctenophore nervous system have been
described, including a subepithelial nerve net (SNN), mesogleal neurons,
tentacular nerves, and elements of nervous system in the aboral organ
(Hernandez-Nicaise, 1984; Hertwig, 1880; Jager et al., 2010). These
neural structures are characterized by unique features such as a
syncytial architecture (Burkhardt et al., 2023) and an extensive
repertoire of lineage-specific neuropeptides not found in other
metazoans (Hayakawa et al., 2022; Sachkova et al., 2021). Notably, early
single-cell transcriptomic studies failed to identify canonical neuronal
clusters using conventional markers (Sebé-Pedrós et al., 2018). It was
only through the use of pro-neuropeptide-based markers that distinct
neural cell types were subsequently identified (Hayakawa et al., 2022;
Sachkova et al., 2021). In parallel, phylogenomic and chromosomal
synteny analyses have suggested that ctenophores may represent the
sister group to all other animals (Li et al., 2021; Ryan et al., 2013;
Schultz et al., 2023; Whelan et al., 2015), raising the possibility that
their nervous system evolved independently from those of other metazoans
(Moroz, 2015). This hypothesis positions ctenophores as a key lineage
for re-evaluating the origins and diversity of neural systems. However,
without a detailed understanding of how ctenophore neurons are spatially
organized and interconnected, it remains unclear how their nervous
system gives rise to behavior. Clarifying this relationship between
neural architecture and behavioral control is essential to uncovering
the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores
control behavior, we focused on a unique sensory structure---the aboral
organ. This gravity-sensing organ contains a statocyst supported by four
mechanosensory balancer cells, which act as pacemakers for comb plate
beating and posture control (HORRIDGE, 1965; Tamm, 1980).
Ultrastructural and immunohistochemical studies have revealed that the
aboral organ contains diverse sensory and neural components, including a
distinct neural structure termed the deep nerve net (Aronova, 1974;
HERNANDEZ-NICAISE, 1968; Jager et al., 2010). A key feature of balancer
cilia is their bidirectional, mechanoresponsive control of beat
frequency. Depending on the direction of deflection and the animal's
geotactic state, balancer cilia may either increase or decrease their
activity, indicating a dynamic response system not fixed like vertebrate
hair cells (HORRIDGE, 1965; Tamm, 1982; Tamm, 2014a). Previous studies
have implicated membrane potential, intracellular calcium concentration,
and synaptic input in modulating this response (Lowe, 1997). In
particular, deflection-induced excitation can be abolished by calcium
channel inhibitors or calcium-free conditions (Lowe, 1997), while high
extracellular K⁺ depolarizes and activates cilia in the absence of
mechanical stimuli (Lowe, 1997). Local calcium application further
suggests that Ca²⁺ influx at the ciliary base may be required for the
excitatory response (Lowe, 1997). These findings support the hypothesis
that neural modulation plays a critical role in shaping balancer cell
output (Lowe, 1997).

To investigate how balancer cilia are modulated within the aboral organ,
we combined volume electron microscopy and high-speed imaging in the
ctenophore \emph{Mnemiopsis leidyi}. Through connectomic reconstruction,
we characterized the synaptic architecture of the aboral nerve net, and
functional imaging revealed correlated activity patterns in balancer
cilia. These results suggest that the aboral nerve net plays a central
role in integrating sensory input and coordinating motor output.

@@

For vEM analysis, we used a five-day-old cydippid larva of
\emph{Mnemiopsis leidyi}. To obtain high-quality ultrastructural
preservation, we used high pressure freezing followed by freeze
substitution and resin embedding (Epon). The specimen was sectioned from
the aboral tip and we collected approximately 1,000 ultra-thin (50 nm)
serial sections as ribbons. The sections were imaged in a Zeiss Gemini
500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only
the region containing the aboral organ. Subsequently, 620 of these
images were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We
skeletonised and annotated all cells in this dataset in CATMAID
(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells with
nuclei and cell bodies intact. Most of the cells, including the
balancers and the aboral nerve net neurons are fully contained within
the volume, due to the compact and mostly self-contained organisation of
the aboral organ.

@@

Next, we divided the entire aboral organ into four broad quadrants
(Q1-Q4) to facilitate grouping the identified cells (Figure 1G)
(Martindale and Henry, 1999). The general body plan of ctenophores, when
viewed from the aboral side, exhibits biradial symmetry around the anal
pores (Martindale and Henry, 1999). This symmetry corresponds to the
four blastomeres present at the four-cell stage during early embryonic
development (Martindale and Henry, 1999). We assigned each of the 909
skeletonised cells to one of the four quadrants.

@@

{\centering \includegraphics{figures/Figure1.png}

@@

\caption{\textbf{Figure 1. Morphology of the aboral organ in
\emph{Mnemiopsis leidyi}}\\
(A) Whole-body image of a 5-day-old \emph{M. leidyi} cydippid larva in
lateral view. The boxed region indicates the aboral organ (ao). Scale
bar: 100 µm.\\
(B) Schematic diagram of the aboral view of a cydippid larva. The
two-tone coloration represents the biradial symmetry of the body. For
convenience, the body is divided into four quadrants, designated as the
first (Q1) through fourth (Q4) quadrants. The two primary viewing angles
are referred to as the sagittal plane (S) and the tentacular plane (T).
Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates;
tb, tentacle bulbs.\\
(C) Aboral views of the aboral organ highlighting its spatial
organization. The left panel presents a schematic representation of the
aboral organ, illustrating the four body quadrants (Q1--Q4), color-coded
consistently with panel (B). The right panel shows a corresponding
differential interference contrast (DIC) image from the same
perspective. The region enclosed by the dotted line delineates the
boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows
indicate the viewing directions corresponding to those planes. Scale
bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\\
(D) Lateral views of the aboral organ in two orthogonal planes captured
using DIC microscopy. The left panel shows a lateral view in the
sagittal (S) plane, while the right panel displays a view in the
tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are
enclosed within the dome (do). Scale bar: 50 µm. A--O indicates the
aboral--oral body axis.\\
(E) Schematic representations of the aboral organ in sagittal and
tentacular planes. The left panel shows a lateral schematic in the
sagittal plane (S), and the right panel shows the same in the tentacular
plane (T), corresponding to the views in panel (D). Colors match the
quadrant scheme shown in panels (B) and (C).\\
(F) Example of cell tracing using the collaborative annotation toolkit
CATMAID for large-scale electron microscopy image datasets. The
spherical objects indicate nuclear positions, while the lines represent
the traced cell centers. Scale bar: 10 µm.\\
(G) Three-dimensional reconstruction of cells composing the aboral
organ, displayed from different perspectives. The left panel presents an
aboral view of the aboral organ. The middle panel shows a sagittal plane
view, while the right panel provides a tentacular plane view. Cells are
color-coded according to their respective quadrants. Lithocytes (li) are
represented as three gray spheres, while balancers (bal) are depicted as
gray lines. Scale bar: 25 µm.}

@@

Classical neural staining techniques did not provide clear images of
neurons at the aboral organ. However, ultrastructural studies revealed
morphological evidence for neural elements, including synapses that are
located on the epithelial floor of the aboral organ---a region where
epithelial cells are densely packed and form a floor-like structure
(Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). In our dataset, we
identified synapses based on the previously described pre-synaptic triad
morphology consisting of a single layer of vesicles, a cisterna of
smooth endoplasmic reticulum and an associated mitochondrion. At
synaptic sites, we marked mitochondria as synaptic nodes (orange in
CATMAID) and connected this node to the nearest node in postsynaptic
cells across the regions where synaptic vesicles aligned with the
presynaptic cell's membrane (light blue arrows in CATMAID)
(Supplementary Figure S2A). We could not detect specialized
post-synaptic structures, in agreement with previous studies
(Hernandez-Nicaise, 1973). Synapses were identified as either monoadic
or polyadic, with one pre-synaptic neuron forming connections with one
or multiple post-synaptic cells (Supplementary Figure S2B).

@@

Furthermore, ANNs are also different from other neurons previously
reported by EM, including mesogleal neurons and ciliated sensory cells
that synapse on the SNN (types 1-4)(Burkhardt et al., 2023). Based on
their distinct size and position, we classified the ANNs into two types.
The first type is a single large neuron (ANN\_Q1-4) spans all four
quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has
150 presynaptic sites. The second type consists of two neurons
(ANN\_Q1Q2 and ANN\_Q3Q4), each spanning two quadrants and containing
two nuclei. ANN\_Q1Q2 had 119 presynaptic structures, whereas ANN\_Q3Q4
had 68 presynaptic structures.

@@

{\centering \includegraphics{figures/Figure2.png}

@@

\caption{\textbf{Figure 2. Organisation of the aboral synaptic nerve
net}\\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4),
which spans all four quadrants and contains multiple nuclei (blue). The
spheres represent the positions of individual nuclei. The left panel
shows a aboral view of the aboral organ, the middle panel presents a
sagittal plane view, and the right panel provides a tentacular plane
view.\\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each
spanning two quadrants and containing multiple nuclei. The spheres
indicate the positions of the nuclei. The left panel shows a aboral view
of the aboral organ, the middle panel presents a sagittal plane view,
and the right panel provides a tentacular plane view.\\
(C) Graph showing the number of mitochondria per cell and the proportion
of mitochondria involved in synapse formation for each cell type. Green
represents mitochondria associated with synapses, while gray represents
mitochondria not involved in synapse formation. ANN has a significantly
higher number of mitochondria compared to other cell types, with most of
them forming synapses. Except for lithocytes, each cell type has a
sample size of at least three cells. The abbreviations are as follows:
bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv
(dense vesicle cells), imc (intra-multiciliated cells), la (lamellate
bodies), li (lithocytes), pl (plumose), and ef (epithelial floor
cells).\\
(D) Localization of mitochondria within ANN Q1-4. Red indicates
mitochondria associated with the presynaptic triad structures, yellow
marks mitochondria containing synaptic vesicles but lacking a clearly
defined presynaptic triad, blue represents mitochondria with unclear
synaptic vesicles, and black denotes mitochondria where no synaptic
vesicles were identified.\\
(E) Representative electron micrographs of presynaptic triad structures
observed in the dataset and their schematic diagrams. The left diagram
illustrates synaptic projections from the center ANN (ANN Q1-4) to the
lateral ANN (ANN Q3Q4), while the right diagram shows synaptic
projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN
Q1-4). The abbreviations are as follows: mi (mitochondrion), er
(endoplasmic reticulum), and sv (synaptic vesicles).\\
(F) Plot showing the three-dimensional positions of synapses. Synapses
from the center ANN to the lateral ANN are shown in magenta, while
synapses from the lateral ANN to the center ANN are shown in cyan. Blue
dots indicate the locations of autapses within ANN Q1-4.}

@@

The second group of cells that form synaptic contact with the ANN were
the bridge cells. Bridge cells, first described by Tamm et al.~in 2002
(Tamm and Tamm, 2002), are characterized by bundles of elongated
processes filled with microtubules that arch over the epithelial layer,
resembling a bridge. Their somata are located at the base of the paired
balancer-cell clusters along the tentacle surface and extend across the
sagittal plane towards the base of the opposite balancer cells. Bridge
cells form two distinct cell groups across the sagittal plane, in the
Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge
cells, in Q3Q4, 12 cells.

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\usepackage{bookmark}

@@

\title{Neural Connectome of the Ctenophore Statocyst}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\textbf{Kei Jokura}\(^{1,2,3,4,5*}\), \textbf{Sanja
Jasek}\(^{1,2,6}\), \textbf{Lara Kewalow}\(^{6}\), \textbf{Pawel
Burkhardt}\(^{7}\), \textbf{Gáspár Jékely}\(^{1,2,6,*}\) \(^{1}\) Living
Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom
\(^{2}\) Biosciences, Faculty of Health and Life Sciences, University of
Exeter, Exeter EX4 4QD, UK \(^{3}\) Grass Laboratory, Marine Biological
Laboratory, Woods Hole, MA 02543, USA \(^{4}\) Exploratory Research
Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan
\(^{5}\) National Institute for Basic Biology (NIBB), Okazaki, 444-8585,
Japan \(^{6}\) Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany \(^{7}\) Michael Sars Centre,
University of Bergen, Norway \textbf{\(^{*}\) Correspondence:}
\includegraphics[width=30em,height=\textheight,keepaspectratio]{pictures/emails.png}}
\author{}
\date{}

@@

Ctenophores possess a unique gravity receptor (statocyst) in their
aboral organ formed by four clusters of ciliated balancer cells that
collectively support a statolyth. During reorientation, differential
load on the balancer cilia leads to altered beating of the ciliated comb
rows to elicit turns. To study the neural bases of gravity sensing, we
imaged by volume electron microscopy (vEM) the aboral organ of the
ctenophore \emph{Mnemiopsis leidyi}. We reconstructed 909 cells,
including syncytial neurons that form a nerve net. The syncytial neurons
synapse on the balancer cells and also form reciprocal connection with
the bridge cells that span the statocyst. High-speed imaging revealed
that balancer cilia beat and arrest in a coordinated manner but with
differences between the sagittal and tentacular planes of the animal,
reflecting nerve-net organisation. Our results suggest a coordinating
rather than sensory-motor function for the nerve net and inform our
understanding of the diversity of nervous-system organisation across
animals.

@@

Ctenophores, or comb jellies, are gelatinous marine animals that
actively swim by beating rows of fused cilia known as comb plates. These
coordinated ciliary movements not only generate propulsion but also
allow precise control of body orientation and directional changes within
the water column. Despite lacking a centralized nervous system,
ctenophores exhibit sophisticated and dynamic behavioural patterns and
respond to a broad range of external stimuli, suggesting the presence of
an evolutionarily unique and functionally integrated sensorimotor
system. Several components of the ctenophore nervous system have been
described, including a subepithelial nerve net (SNN), mesogleal neurons,
tentacular nerves, and elements of nervous system in the aboral organ
(Hernandez-Nicaise, 1984; Hertwig, 1880; Jager et al., 2010). These
neural structures are characterized by unique features such as a
syncytial architecture (Burkhardt et al., 2023) and an extensive
repertoire of lineage-specific neuropeptides not found in other
metazoans (Hayakawa et al., 2022; Sachkova et al., 2021). Notably, early
genomic and single-cell transcriptomic studies failed to identify
conserved neuronal markers in ctenophores (Moroz et al., 2014;
Sebé-Pedrós et al., 2018). It was only through the use of
pro-neuropeptide-based markers that distinct neural cell types were
molecularly identified (Hayakawa et al., 2022; Sachkova et al., 2021).
In parallel, phylogenomic and chromosomal synteny analyses have
suggested that ctenophores may represent the sister group to all other
animals (Li et al., 2021; Ryan et al., 2013; Schultz et al., 2023;
Whelan et al., 2015), raising the possibility that their nervous system
evolved independently from those of other metazoans (Moroz, 2015). This
hypothesis positions ctenophores as a key lineage for re-evaluating the
origins and diversity of neural systems. However, without a detailed
understanding of how ctenophore neurons are spatially organized and
interconnected, it remains unclear how their nervous system controls
behaviour. Clarifying this relationship between neural architecture and
behavioural control is essential to uncovering the logic of ctenophore
neural function.

To address the lack of structural, circuit-level understanding of the
ctenophore nervous system, we focused on a unique sensory
structure---the aboral organ. This gravity-sensing organ contains a
statocyst supported by four mechanosensory balancer cells, which act as
pacemakers for the beating of comb plate cilia and thereby control body
orientation (Horridge, 1965; Tamm, 1980). Ultrastructural and
immunohistochemical studies have revealed that the aboral organ contains
diverse sensory and neural components, including a distinct neural
structure termed the deep nerve net (Aronova, 1974; Hernandez-Nicaise,
1968; Jager et al., 2010). A key feature of balancer cilia is their
bidirectional, mechanoresponsive control of beat frequency. Depending on
the direction of deflection and the animal's geotactic state, balancer
cilia may either increase or decrease their activity, indicating a
dynamic response system, not a fixed one like in vertebrate hair cells
(Horridge, 1965; Tamm, 1982; Tamm, 2014a). Previous studies have
implicated membrane potential, intracellular calcium concentration, and
synaptic input in modulating this response (Lowe, 1997). In particular,
deflection-induced excitation can be abolished by calcium channel
inhibitors or calcium-free conditions (Lowe, 1997), while high
extracellular K⁺ depolarizes and activates cilia in the absence of
mechanical stimuli (Lowe, 1997). Local calcium application further
suggests that Ca²⁺ influx at the ciliary base may be required for the
excitatory response (Lowe, 1997). These findings support the hypothesis
that neural modulation plays a critical role in shaping balancer-cell
output (Lowe, 1997).

To investigate how balancer cilia are modulated within the aboral organ,
we combined volume electron microscopy and high-speed imaging in the
ctenophore \emph{Mnemiopsis leidyi}. Through connectomic reconstruction,
we characterized the synaptic architecture of the aboral nerve net,
revealing synaptic inputs to the balancer cell. By functional imaging,
we uncovered correlated activity patterns in balancer cilia across the
four quandrants of the apical organ. Our results suggest that the aboral
nerve net functions in coordinating the activity of balancer cilia,
rather than fulfilling a feedforward sensory-motor function.

@@

For vEM analysis, we used a five-day-old cydippid larva of
\emph{Mnemiopsis leidyi}. To obtain high-quality ultrastructural
preservation, we used high pressure freezing followed by freeze
substitution and resin embedding (Epon). The specimen was sectioned from
the aboral tip and we collected approximately 1,000 ultra-thin (50 nm)
serial sections as ribbons. The sections were imaged in a Zeiss Gemini
500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only
the region containing the aboral organ. Subsequently, images from 620
layers were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We
skeletonised and annotated all cells in this dataset in CATMAID
(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells that
had intact nuclei and cell bodies. Most of the cells, including the
balancers and the aboral nerve net neurons are fully contained within
the volume, due to the compact and mostly self-contained organisation of
the aboral organ.

@@

Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to
facilitate grouping the identified cells (Figure 1G) (Martindale and
Henry, 1999). The general body plan of ctenophores, when viewed from the
aboral side, exhibits biradial symmetry around the anal pores
(Martindale and Henry, 1999). This symmetry corresponds to the four
blastomeres present at the four-cell stage during early embryonic
development (Martindale and Henry, 1999). We assigned each of the 909
skeletonised cells to one of the four quadrants.

@@

{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure1.png}}

@@

\caption{\textbf{Figure 1. Morphology of the aboral organ in
\emph{Mnemiopsis leidyi}}\\
(A) Whole-body image of a 5-day-old \emph{M. leidyi} cydippid larva in
lateral view. The boxed region indicates the aboral organ (ao). Scale
bar: 100 µm.\\
(B) Schematic diagram of the aboral view of a cydippid larva. The
two-tone coloration represents the biradial symmetry of the body. For
convenience, the body is divided into four quadrants, designated as the
first (Q1) through fourth (Q4) quadrants. The two primary viewing angles
are referred to as the sagittal plane (S) and the tentacular plane (T).
Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates;
tb, tentacle bulbs.\\
(C) Aboral views of the aboral organ highlighting its spatial
organization. The left panel presents a schematic representation of the
aboral organ, illustrating the four body quadrants (Q1--Q4), color-coded
consistently with panel (B). The right panel shows a corresponding
differential interference contrast (DIC) image from the same
perspective. The region enclosed by the dotted line delineates the
boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows
indicate the viewing directions corresponding to those planes. Scale
bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\\
(D) Lateral views of the aboral organ in two orthogonal planes captured
using DIC microscopy. The left panel shows a lateral view in the
sagittal (S) plane, while the right panel displays a view in the
tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are
enclosed within the dome (do). Scale bar: 50 µm. A--O indicates the
aboral--oral body axis.\\
(E) Schematic representations of the aboral organ in sagittal and
tentacular planes. The left panel shows a lateral schematic in the
sagittal plane (S), and the right panel shows the same in the tentacular
plane (T), corresponding to the views in panel (D). Colors match the
quadrant scheme shown in panels (B) and (C).\\
(F) Example of cell tracing in CATMAID. The spherical objects indicate
nuclear positions, while the lines represent the traced cell centers.
Scale bar: 10 µm.\\
(G) Morphological rendering of cells composing the aboral organ
displayed in aboral view (left pannel), sagittal plane view (middle
panel) and tentacular plane view (right panel). Cells are color-coded
according to quadrants. Lithocytes (li) are represented as three gray
spheres, balancers (bal) are depicted as gray lines. Scale bar: 25 µm.}

@@

Classical neural staining techniques did not provide clear images of
neurons at the aboral organ. However, ultrastructural studies revealed
morphological evidence for neural elements, including synapses that are
located on the epithelial floor of the aboral organ---a region where
epithelial cells are densely packed and form a floor-like structure
(Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). In our dataset, we
identified 428 synapses based on the previously described pre-synaptic
triad morphology consisting of a single layer of vesicles, a cisterna of
smooth endoplasmic reticulum and an associated mitochondrion. At
synaptic sites, we marked mitochondria as synaptic nodes (orange in
CATMAID) and connected this node to the nearest node in postsynaptic
cells across the regions where synaptic vesicles aligned with the
presynaptic cell's membrane (light blue arrows in CATMAID) (Figure
2--figure supplement 1A). We could not detect specialized post-synaptic
structures, in agreement with previous studies (Hernandez-Nicaise,
1973). Synapses were identified as either monoadic or polyadic, with one
pre-synaptic neuron forming connections with one or multiple
post-synaptic cells (Figure 2--figure supplement 1B).

@@

Furthermore, ANNs are also different from other neurons previously
reported by EM, including mesogleal neurons and ciliated sensory cells
that synapse on the SNN (types 1-4)(Burkhardt et al., 2023). Based on
their distinct size and position, we classified the ANNs into two types.
The first type is a single large neuron (ANN\_Q1-4) that spans all four
quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has
150 presynaptic sites. The second type consists of two neurons
(ANN\_Q1Q2 and ANN\_Q3Q4), each spanning two quadrants and containing
two nuclei. ANN\_Q1Q2 had 119 presynaptic structures, whereas ANN\_Q3Q4
had 68 presynaptic structures.

@@

{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure2.png}}

@@

\caption{\textbf{Figure 2. Organisation of the aboral synaptic nerve
net}\\
(A) Morphological rendering of the large aboral nerve-net neuron
ANN\_Q1-4 spanning all four quadrants, in aboral view (left), sagittal
view (middle panel) and tentacular view (right panel). Spheres indicate
the positions of nuclei.\\
(B) Morphological rendering of the two smaller ANN neurons ANN\_Q1Q2
(pink) and ANN\_Q3Q4 (orange), each spanning two quadrants, in aboral
view (left), sagittal view (middle panel) and tentacular view (right
panel). Spheres indicate the positions of nuclei.\\
(C) Number of mitochondria per cell for each cell type. The number of
synapse-associated mitochondria is shown in green, the number of
mitochondia outside synapses is shown in grey. N \textgreater{} 2 cells
for each cell type except lithocytes. Abbreviations: bal, balancer; brg,
bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc,
intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl,
plumose; ef, epithelial floor cells.\\
(D) Position of mitochondria within ANN\_Q1-4. Red indicates
mitochondria associated with the presynaptic triad structures, yellow
marks mitochondria containing synaptic vesicles but lacking a clearly
defined presynaptic triad, blue represents mitochondria with unclear
synaptic vesicles, and black denotes mitochondria where no synaptic
vesicles were identified.\\
(E) Representative electron micrographs of presynaptic triad structures
observed in the dataset and their schematic diagrams. The left diagram
illustrates synaptic projections from the center ANN (ANN Q1-4) to the
lateral ANN (ANN Q3Q4), while the right diagram shows synaptic
projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN
Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv,
synaptic vesicles.\\
(F) Position of synapses. Synapses from the center ANN to the lateral
ANN are shown in magenta, while synapses from the lateral ANN to the
center ANN are shown in cyan. Blue dots indicate the locations of
autapses within ANN\_Q1-4.}

@@

The second group of cells that formed synaptic contacts with the ANN
were the bridge cells. Bridge cells, first described in 2002 (Tamm and
Tamm, 2002), are characterized by bundles of elongated processes filled
with microtubules that arch over the epithelial layer, resembling a
bridge. Their somata are located at the base of the paired balancer-cell
clusters along the tentacle surface and extend across the sagittal plane
towards the base of the opposite balancer cells. Bridge cells form two
distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4
regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12
cells.
